Climate-driven phytoplankton community variability in the Cariaco Basin, Venezuela

### 2025-06-25 agujournaltemplate.tex
post-Ago Review working at Methods

\begin{keypoints}
\item The CARIACO time series phytoplankton community data reveals a shift and return between two clusters, defined by high- and low-upwelling conditions that correspond to anomalies in the AMO index. 
\index Between the clusters, we observe no significant changes in diversity metrics, except for lower turnover and higher community evenness under low-upwelling conditions.
\item Gradient forest analysis identified the AMO climate index as the strongest predictor of community changes, highlighting the influence of large-scale climatic oscillations on the tropical coastal ecosystem.
\end{keypoints}

@@

\begin{abstract}
Phytoplankton play a crucial role in Earth's biogeochemical cycles. Climate drives changes in the physical environment of phytoplankton communities, which can, in turn, affect the productivity and diversity of the entire ecosystem. Long-term time series can capture changes in climate and the ecosystem, allowing for the untangling of effects acting on a local population. 
This study aims to describe and investigate the drivers of changes observed in the phytoplankton community of the Cariaco Basin, Venezuela, between 1995 and 2017. Monthly cruises sampled microscopy cell counts and biogeochemical properties, which we extend with data from AMO and MEI v.2 climate indices, as well as the ERA5 climate model reanalysis. 
We describe and relate changes in bottom-up drivers of the upwelling system with the time series of phytoplankton biomass and diversity. Using clustering algorithms, we investigate the structure in the community time series data, revealing two clusters that describe a shift from high to low upwelling conditions in 2004, but also show a return to a community and conditions similar to those before the shift in 2014. Our results suggest that community diversity did not follow the same pattern but instead was diminished between 1998 and 2009. To quantify the effect of climate and bottom-up drivers on phytoplankton community changes, we employed a gradient forest analysis and found that a combination of in situ upwelling-related metrics (SST and nitrate concentration) and climate indices (AMO and MEI v.2) were the strongest predictors. Our study emphasizes the impact of decadal-scale climate oscillations, such as the AMO, on the phytoplankton community in a tropical coastal ecosystem through modulation of the upwelling system and highlights the resilience of ecosystems to recover from climate changes.

@@

    % Phytoplankton is important
    Phytoplankton provide the foundation for marine food webs and play a crucial role in biogeochemical processes, including oxygen production and carbon export to the deep ocean \cite{falkowski_biogeochemical_1998}. The paraphyletic group consists of a diverse set of organisms that exhibit highly variable ecophysiological properties \cite{appeltans_magnitude_2012}. The short generation time and high diversity allow phytoplankton communities to adapt quickly to environmental changes, making them suitable indicators of ecosystem state \cite{alvarez-cobelas_what_1998, barton_anthropogenic_2016, di_cavalho_temporal_2023}.
    Coastal margins constitute \qty{\sim 10}{\%} of the total ocean surface area but are responsible for an estimated \qty{40}{\%} of carbon sequestration to the seafloor \cite{yool_examination_2001, mullerkarger_importance_2005}. Using remote sensing, a global increase in coastal phytoplankton bloom frequency and spatial extent has been observed in the last 20 years, although regional effects are highly variable \cite{dai_coastal_2023}.
    Climate change is predicted to have a lasting impact on phytoplankton communities, shifting the ranges and contributions of functional groups \cite{boyd_biological_2016, henson_future_2021}. 
    Long-term time series of biological observations are crucial for understanding how marine ecosystems respond to temporary shifts, cyclic processes, and long-term changes in the local physical environment and climate \cite{carstensen_need_2014, henson_observing_2016}. Such time series are rare, particularly in the tropics \cite{clarke_does_2017}, where community diversity is often highest \cite{brown_why_2014, righetti_global_2019} and faces greater threats of extinction compared to temperate regions \cite{finnegan_paleontological_2015}.

    A tropical coastal ecosystem that has been studied intensively is the CARIACO time series, which has involved consistent sampling of physical and biogeochemical parameters in the Cariaco Basin from 1995 to 2017 \cite{muller-karger_scientific_2019}. 
    % Describe CARIACO location
    The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within the largest truly marine anoxic basin in the world \cite{edgcomb_accessing_2011}. The basin is a continental shelf depression consisting of two sub-basins of \qty{\sim 1400}{m} depths, which are effectively separated from the Caribbean Sea by a sill with a mean depth of about \qty{100}{m}. The water below the sill is poorly ventilated, which, combined with the high surface productivity and microbial respiration of sinking organic matter, results in permanent anoxic conditions at depths below \qty{\sim 275}{m} \cite{thunell_organic_2000}. Due to these unique properties, the Cariaco Basin has been used as a natural laboratory since the 1950s, providing a rich sedimentary record for the study of past climate \cite{hughen1996nature} and serving as a testbed for studying the coupling between pelagic production and carbon export to the deep ocean \cite{montes_vertical_2012}.
    % Describe general seasonal cycle with functional groups etc.
    A markedly seasonal upwelling cycle is driven by easterly trade winds, with the highest levels of primary production observed from December through May when the Inter-Tropical Convergence Zone (ITCZ) moves southward. A smaller wind-driven secondary upwelling often occurs in July and August \cite{mullerkarger_annual_2001, astor_seasonal_2003}. During the upwelling period, the supply of deeper nutrient-rich waters supports high biological productivity, and the phytoplankton community is dominated by diatoms \cite{romero_seasonal_2009}, with dinoflagellates, coccolithophorids, and nanoplankton also contributing. In the low-wind rainy season, the Cariaco Basin exhibits oligotrophic conditions, where the fraction of large phytoplankton cells is diminished, but nano- and picophytoplankton are present in considerable numbers throughout the year \cite{lorenzoni_characterization_2015}.

    % PARAGRAPH THREE - CHANGES IN CARIACO
    Since the inception of the time series in 1995, the ecosystem of the Cariaco Basin has undergone marked changes. \citeA{taylor_ecosystem_2012} analyzed time series data up to 2010 and found significant decadal trends of increasing sea surface temperature (SST), coinciding with a decrease in upwelling and phytoplankton bloom intensities, which the authors linked to the impact of global warming. During the initial phase of the time series (1996–2004), the Cariaco Basin underwent a consistent seasonal upwelling cycle, driven by strong trade winds during winter and spring (December – April) \cite{mullerkarger_annual_2001, astor_seasonal_2003}. Following 2004, there was a transition in the physical and biogeochemical regime characterized by a significant reduction in trade winds during the upwelling period \cite{taylor_ecosystem_2012}. This strongly affected the ecosystem and phytoplankton community. Microscopy cell counts indicate the collapse of large phytoplankton, particularly diatoms, as confirmed to a lesser extent by pigment data analysis \cite{pinckney_phytoplankton_2015}. In general, a shift was observed in the phytoplankton community toward smaller cell sizes, along with increased abundances of coccolithophores, cryptophytes, and other phytoflagellates, which are typical of stratified water columns \cite{lorenzoni_characterization_2015}. An analysis of the pigment data indicates a deepening of the chlorophyll maximum, an increase in diversity, and a reduction in seasonal variability in community composition \cite{pinckney_phytoplankton_2015}. During this phase, the heavily exploited sardine fisheries in coastal Venezuela experienced a collapse, coinciding with an increase in zooplankton biomass measured in the Cariaco Basin \cite{taylor_ecosystem_2012}. Interestingly, changes in the phytoplankton community did not result in a significant change in the export of particulate organic carbon (POC) to depth, as measured by sediment traps deployed during the time series \cite{pinckney_phytoplankton_2015}.

    % PARAGRAPH FOUR - EXPLANATION OF CHANGES // OPEN QUESTIONS!!
    The shift in the community, due to the warming trend and the reduction in upwelling intensities, does not appear to be a permanent regime shift. Data from 2012 until the end of the CARIACO time series in 2017 indicate a return to stronger upwelling intensities and increased numbers of diatoms and dinoflagellates \cite{muller-karger_scientific_2019}. This return in 2014 to conditions similar to those of the late 1990s coincided with a recovery in sardine catches observed in 2016 and 2017 in a location southeast of Isla de Margarita \cite{gomez_gaspar_variacion_2025}. 
    The impact of large-scale climatic processes on the observed shifts in the community remains unclear. The Atlantic Multidecadal Oscillation (AMO) has emerged as an important indicator of the climate system and its effects on the ecosystems of the north Atlantic \cite{nye_ecosystem_2014}. The oscillations of the AMO index are strongly correlated with temperature indices in the Caribbean region \cite{stephenson_changes_2014}.
    The El Niño Southern Oscillation (ENSO) has a teleconnection to the Caribbean Sea, as during ENSO (indicated by a positive MEI v.2 index) sea surface temperatures rise and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}. However, \citeA{taylor_ecosystem_2012} found no strong correlation between the MEI v.2 and measurements in the Cariaco Basin and did not test for correlations with the AMO index.

    % What drives the changes??? Bottom-up/ Top-down ... CLIMATE 
    \citeA{taylor_ecosystem_2012} cautioned that the observed changes cannot be clearly identified as "unidirectional trends driven by anthropogenic climate change or whether they simply reflect low-frequency natural cycles, such as those driving the Atlantic Multidecadal Oscillation."
    Previous analyses of phytoplankton community diversity did not include time series data beyond 2011, and so far there was no detailed investigation of changes in diversity based on the microscopy cell count data.
    From this background, we aim to 1) investigate changes in the depth distribution of phytoplankton biomass and community diversity over the entire time series, 2) relate the structure of the phytoplankton community data to the shifts observed in physical and biogeochemical variables, and 3) assess how decadal-scale climate osciallations as signified by the ENSO MEI v.2 and AMO index relate to changes in the phytoplankton community.
    Our analysis focuses exclusively on bottom-up drivers of community changes, as these data are available throughout the entire time series and at a monthly resolution. This study complements the work of \citeA{taylor_ecosystem_2012} and \citeA{pinckney_phytoplankton_2015} with a more detailed analysis of biodiversity aspects using all data available from the CARIACO time series up to the final cruise in 2017. Understanding the regional dynamics of the Cariaco Basin is crucial for comprehending the broader ecological changes in tropical marine ecosystems driven by shifts in climate.

\section{Materials and Methods}
% Here is text on Materials and Methods.
%
\subsection{Study Site}
    %Talk about the Basin shortly, then about the time series
    The Cariaco Ocean Time Series Program was established to explore the relationship between ocean surface processes and the sinking flux of particulate carbon. The Cariaco Basin is a large (approximately 160 km long and 70 km wide) and deep (approximately 1,400 m) basin, situated on the northeastern Venezuelan continental shelf, with a sill to the north at a mean depth of around \qty{100}{m}. This area is relatively productive ($\sim$ \qty{400}{g.C.m^{-2}.y^{-1}}), with primary production and vertical particulate organic matter fluxes exceeding those observed at BATS and HOT sites, resulting in the accumulation of anoxic bottom waters below \qty{\sim 200}{m} depth.

\subsection{In-Situ Data}
    Samples were collected as part of the Cariaco Ocean Time Series Program during monthly research vessel cruises to a station located in the eastern Cariaco Basin (\ang{10.50}N, \ang{64.66}W) \cite{muller-karger_scientific_2019}.
    The CARIACO field program consisted of a monthly core oceanographic cruise with Niskin bottle and CTD deployments, in addition to other deployments and samples not used in this manuscript. Field logistics and partial sample analysis were coordinated by the Fundación La Salle de Ciencias Naturales on Margarita Island, Venezuela. Details of the history of sampling procedures in CARIACO, as well as sampling schemes, analytical methods, and data quality control and inter-calibration procedures, have been described by \citeA{mullerkarger_annual_2001, thunell_particulate_2007, astor_yrene_m_handbook_2013, astor_interannual_2013, astor_sintesis_2014}.

@@

    Between late 1995 and early 1998, nutrient data was provided by William Senior (Universidad de Oriente, Cumana, Venezuela) and from mid-1997 to the present by Dr. K. Fanning (University of South Florida, St. Petersburg, FL). Samples from eight cruises were analyzed by both laboratories. Data for $PO_4$ and $SiO_4$ were in strong agreement (within ±5\%), whereas the $NO_x$ and $NH_4$ measurements showed discrepancies. \cite{taylor_ecosystem_2012}. For our analysis, we used a merged dataset for all nutrients, where coincidental monthly measurements were averaged.

@@

    Data sets for Niskin bottle samples, CTD, and phytoplankton taxonomy were retrieved from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) \url{https://www.bco-dmo.org/project/2047} (accessed in February, 2024). 
    From the CARIACO Time Series data, we used only those variables that cover the entire length of the time series without large gaps (i.e. no more than three months) in coverage. Therefore, we did not include zooplankton net tow and HPLC pigment data in our analysis. Our study instead focused on the phytoplankton microscopy data and bottom-up drivers of change. We compared these data with Niskin bottle nutrient, temperature, and salinity data, as well as climate data and indices, focusing on bottom-up drivers of community variability and diversity. In total, we used a complete set of 254 monthly data points, on which all variables were available.

\subsection{Climate data}
    %from Taylor et al. 2012
    Zonal wind stress (-u, easterly Trade Winds) across the southern Caribbean Sea drives nearshore upwelling of nutrient-rich waters via Ekman transport \cite{rueda-roa_southern_2013}. We extracted the -u wind component at 10 m from the ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis in the grid cell closest to the Cariaco time series sampling coordinates (\ang{10.4}N, \ang{64.8}W, 31 km grid size), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels} (accessed June 2024). This variable is referred to as "wind speed" throughout the manuscript for simplicity. The data for precipitation and evaporation are similarly extracted from the ERA5 model reanalysis dataset for the aforementioned grid cell. Evaporation provides an estimate of the accumulated amount of water that has evaporated from the Earth's surface, while precipitation is the estimate of total precipitation, comprising all liquid that falls to Earth's surface as rain or snow.

    The monthly AMO index data (10-year low-pass), which represents an area average of detrended low-pass filtered North Atlantic SST anomalies, were obtained from \url{https://climatedataguide.ucar.edu/climate-data/atlantic-multi-decadal-oscillation-amo} (accessed in June, 2024).

    The bi-monthly multivariate El Niño/Southern Oscillation (ENSO) indices (MEI v.2), defined as the time series of the leading combined Empirical Orthogonal Function (EOF) of five different variables (sea level pressure (SLP), sea surface temperature (SST), zonal and meridional components of surface wind, and outgoing longwave radiation (OLR)) over the tropical Pacific basin (\ang{30}S-\ang{30}N and \ang{100}E-\ang{70}W), were obtained from \url{https://psl.noaa.gov/enso/mei/} (accessed in June, 2024). In our analysis, bi-monthly values are aligned with the tail month (December-January is treated as the value for January).

    \subsubsection{Data processing}
    % Niskin, CTD and Phytoplankton data processing
    Data were processed and analyzed using the R programming language (version 4.3.3) \cite{r_core_team_r_2024} with data pipelines created using tidyverse packages \cite{wickham_welcome_2019}. All data processing and analysis code is publicly available on GitHub \url{https://github.com/ben1post/PhytoCariaco}.
    For all discrete depth measurements, the data were interpolated using the "oce" package \cite{kelley_oce_2023} with the adaptive "unesco" algorithm defined by the U.S. National Oceanographic Data Center \cite{johnson2006world}. The interpolated values were depth integrated to 100 meters and in four discrete intervals between 0-25, 25-50, 50-75 and 75-100 \meter for further analysis. Additionally, sea surface temperature (SST) was calculated from CTD temperature measurements averaged over the top \qty{10}{\meter}. For phytoplankton microscopy data, individually identified taxonomic units (species or genus) were interpolated and integrated across depth, and finally grouped by genus via the sum of interpolated counts.

\subsection{Data analysis}    
    To provide a general overview of the data and trends in the time series, we calculated annual means for relevant variables and normalized the data range as z-scores ($z = \frac{(x – \mu)}{\sigma}$). Cruises from 1995, 2016, and 2017 were excluded due to insufficient measurements to provide a reliable yearly mean estimate.

@@

    \subsubsection{Phytoplankton Community Structure Analysis}
    To investigate the structure of the phytoplankton time series data, we used a complete linkage clustering method via the "hclust" function in R \cite{r_core_team_r_2024} on the binary Jaccard distance matrix of annual aggregates of observed genera integrated over 100 m depth. Additionally, we performed non-metric multidimensional scaling (NMDS) ordination and hierarchical clustering on the monthly integrated counts that were square-root transformed using "metaMDS" from the "vegan" package \cite{oksanen_vegan_2024}. We employed the Bray distance matrix in a two-dimensional ordination that reached a converging solution within 30 steps. To relate the multidimensional structuring of the community data to variability in environmental data, we used the "envfit" function from the "vegan" package to fit the environmental vectors onto the ordination with 999 permutations. The arrows indicate the gradient of a two-dimensional linear trend surface. 
    To contrast the environmental parameters and diversity data between the community clusters resulting from the yearly aggregated clustering, we created density distribution plots contrasting the monthly measurements between clusters using the "ggpubr" package \cite{kassambara_ggpubr_2023}. To test whether the environmental variables, chlorophyll $a$ and diversity metrics varied significantly between the two community clusters, we employed a two-sided Wilcoxon sum rank test using the "wilcox.test" function in R \cite{r_core_team_r_2024}.
    % detailed explanation of NMDS in Winder and Hunter 2008

>>>>

\begin{keypoints}
\item The CARIACO phytoplankton community time series reveals two clusters associated to high- and low-upwelling conditions and corresponding to anomalies in the AMO index. 
\index No significant differences are observed in diversity metrics between the two clusters, with the exception of a lower turnover and a higher community evenness during low-upwelling conditions.
\item Gradient forest analysis identifies the AMO climate index as the strongest predictor of community change, highlighting the influence of large-scale climatic oscillations on this tropical coastal ecosystem.
\end{keypoints}

@@

\begin{abstract}
Phytoplankton communities play a crucial role in oceans' biogeochemical cycles. Climate drives changes in the physical environment that phytoplankton communities experience and can therefore impact the productivity and diversity of marine ecosystems. Long-term time series can capture changes in climate and ecosystems, allowing us to unravel the effects on local communities. 
This study aims to describe and investigate the drivers of changes in  phytoplankton community data (biomass and diversity) collected in the Cariaco Basin, Venezuela, between 1995 and 2017. The monthly cruises sampled water for microscopy cell counts and determination of biogeochemical properties, which we integrated with data from AMO and MEI v.2 climate indices and ERA5 climate model reanalysis. 
Using clustering algorithms, we investigate the structure of the community data, revealing two temporal clusters associated with high and low upwelling conditions (from 1995 to 2004 and from 2005 to 2017, respectively) and showing a return to community biomass and physical conditions similar to those characterizing the period 1995-2004. However,  community diversity did not follow the same pattern, showing a decrease between 1998 and 2009. To quantify the effect of climate and bottom-up drivers on phytoplankton community changes, we employed a gradient forest analysis and found that a combination of in situ, short-term, upwelling-related variables (SST and nitrate concentration) and long-term, climate indices (AMO and MEI v.2) were the strongest predictors. Our study emphasizes the impact of long-term (decadal) climatic oscillations, such as the AMO, on the phytoplankton community of a tropical coastal ecosystem through modulation of the upwelling system and highlights the capability of the ecosystem to recover from climate change.

@@

    % Phytoplankton is important
    Phytoplankton provide the foundation for marine food webs and play a crucial role in biogeochemical processes, including oxygen production and carbon export to the deep ocean \cite{falkowski_biogeochemical_1998}. This paraphyletic community consists of a diverse group of organisms  exhibiting highly variable eco-physiological properties \cite{appeltans_magnitude_2012}. The short generation time and high diversity allow phytoplankton communities to respond relatively quickly to environmental changes, making them suitable indicators of ecosystem state \cite{alvarez-cobelas_what_1998, barton_anthropogenic_2016, di_cavalho_temporal_2023}.
    Coastal margins, where phytoplankton typically thrive, constitute only \qty{\sim 10}{\%} of the global ocean surface  but are responsible for an estimated \qty{40}{\%} of carbon sequestration to the seafloor \cite{yool_examination_2001, mullerkarger_importance_2005}.

    Data from remote sensing, suggest a global increase in the frequency and spatial extent of coastal phytoplankton blooms over the last 20 years, but regional effects are highly variable \cite{dai_coastal_2023} and difficult to capture.
    Climate change is predicted to have a lasting impact on phytoplankton communities, including long-term shifts in community composition and alterations of diversity \cite{acevedo-trejos_glimpse_2014, boyd_biological_2016, henson_future_2021}. 
    Long-term time series of biological observations are crucial for understanding how marine ecosystems respond to short-term cyclic processes and to long-term changes in the local physical environment and climate \cite{carstensen_need_2014, henson_observing_2016}. Such time series are rare, particularly in the tropics \cite{clarke_does_2017}, where community diversity is often highest \cite{brown_why_2014, righetti_global_2019} and faces greater threats of extinction compared to temperate regions \cite{finnegan_paleontological_2015}.

    A tropical coastal location that has been studied intensively in the last decades is the CARIACO basin, where a wide range of physical and biogeochemical parameters have been consistently collected from 1995 to 2017 \cite{muller-karger_scientific_2019}. 
    % Describe CARIACO location
    The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within one of the largest anoxic basins in the world oceans \cite{edgcomb_accessing_2011}. The basin is a continental shelf depression comprising two sub-basins of \qty{\sim 1400}{m} depths, which are effectively separated from the Caribbean Sea by a sill with a mean depth of about \qty{100}{m}. The water below the sill is poorly ventilated. This underwater topography in combination with a high surface productivity, associated to aerobic degradation of fast sinking organic matter, creates permanent anoxic conditions at depths below \qty{\sim 275}{m} \cite{thunell_organic_2000}. Due to these unique properties, the Cariaco Basin has been used as a natural laboratory since the 1950s, providing a rich sedimentary record for the study of past climate \cite{hughen1996nature} and serving as a testbed for understanding the coupling between pelagic production and carbon export to the deep ocean \cite{montes_vertical_2012}.

    % Describe general seasonal cycle with functional groups etc.
    The Cariaco Basin is characterised by a marked seasonal cycle of upwelling, which is driven by easterly trade winds, with the highest levels of primary production observed from December through May when the Inter-Tropical Convergence Zone (ITCZ) moves southward. A smaller wind-driven secondary upwelling event occurs often in July and August \cite{mullerkarger_annual_2001, astor_seasonal_2003}. During the upwelling period, the supply of deeper nutrient-rich waters supports high biological productivity, and the phytoplankton community is dominated by diatoms \cite{romero_seasonal_2009}, with dinoflagellates, coccolithophorids, and nanoplankton also contributing. In the low-wind, rainy season, the Cariaco Basin exhibits oligotrophic conditions and the fraction of large phytoplankton cells is reduced, albeit nano- and picophytoplankton cells are present in relatively high numbers throughout the year \cite{lorenzoni_characterization_2015}.

    % PARAGRAPH THREE - CHANGES IN CARIACO
    Since the inception of the time series in 1995, the ecosystem of the Cariaco Basin has undergone marked changes. \citeA{taylor_ecosystem_2012} analyzed the time series data up to 2010 and found significant decadal trends of increasing sea surface temperature (SST), coinciding with a decrease in upwelling and phytoplankton bloom intensities, which the authors attributed to global warming. During the initial phase of the time series (1996–2004), the Cariaco Basin underwent consistent and pronounced seasonal upwelling cycles, driven by strong trade winds during winter and spring (December – April) \cite{mullerkarger_annual_2001, astor_seasonal_2003}. After 2004, a significant reduction in the strength of trade winds during the upwelling period \cite{taylor_ecosystem_2012} caused a regime change in the physical and biogeochemical system. This strongly affected the phytoplankton community and the whole ecosystem. Microscopy cell counts indicated a drastic reduction of large phytoplankton, particularly diatoms, as confirmed, albeit not with the same high significance, by analysis on pigment data  \cite{pinckney_phytoplankton_2015}. In general, a shift was observed in the phytoplankton community toward smaller cell sizes, along with increased abundances of coccolithophores, cryptophytes, and other phytoflagellates, groups commonly occurring in stratified water columns \cite{lorenzoni_characterization_2015}. An analysis of pigment data indicates a deepening of the chlorophyll maximum, an increase in diversity, and a reduction in seasonal variability of community composition \cite{pinckney_phytoplankton_2015}. During this phase, the heavily exploited sardine fisheries in coastal Venezuela experienced a collapse, coinciding with an increase in zooplankton biomass \cite{taylor_ecosystem_2012}. Interestingly, changes in the phytoplankton community did not result in a significant change in the export of particulate organic carbon (POC) to depth, as measured by sediment traps deployed during the time series \cite{pinckney_phytoplankton_2015}.

    % PARAGRAPH FOUR - EXPLANATION OF CHANGES // OPEN QUESTIONS!!
    The shift in phytoplankton community composition associated to increased temperature and to reduced upwelling intensities, does not appear to be a permanent change in regime. Data from 2012 to 2017, the end phase of the CARIACO time series, indicates a return to strong upwelling intensities and to high numbers of diatoms and dinoflagellates \cite{muller-karger_scientific_2019}. The return to conditions similar to those of the late 1990s coincides with a recovery in sardine catches observed in 2016 and 2017 in an area southeast of Isla de Margarita \cite{gomez_gaspar_variacion_2025}. 
    The impact of large-scale climatic processes on the observed shifts in the phytoplankton community remains unclear. The Atlantic Multidecadal Oscillation (AMO) has emerged as an important factor influencing the climate system in the region and the ecosystems of the north Atlantic \cite{nye_ecosystem_2014}. The oscillations of the AMO index are strongly correlated with temperature indices in the Caribbean region \cite{stephenson_changes_2014}.
    The El Niño Southern Oscillation (ENSO) is teleconnected to the Caribbean Sea given that during ENSO (indicated by a positive MEI v.2 index) sea surface temperatures increase and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}. However, using the time-series available only until 2010, \citeA{taylor_ecosystem_2012} found no significant correlation between MEI v.2 and in-situ observations. In addition, the AMO index was included in the analyses of the Cariaco Basin.

    % What drives the changes??? Bottom-up/ Top-down ... CLIMATE 
    \citeA{taylor_ecosystem_2012} cautioned that the observed changes cannot be clearly identified as "unidirectional trends driven by anthropogenic climate change", and indicated that they may "simply reflect low-frequency natural cycles, such as those driving the Atlantic Multidecadal Oscillation".
    Previous analyses of phytoplankton community diversity in the Cariaco Basin did not include time series data beyond 2011, and no detailed investigations have been undertaken yet on possible changes in diversity based on microscopy cell counts.
    Given these knowledge gaps, we aim to investigate 1) changes in the depth distribution of phytoplankton biomass and community diversity over the period 1995\,--\,2017, 2) changes in the structure of the phytoplankton community data and potential connections with shifts observed in physical and biogeochemical variables, and 3) potential connections between long-term climatic oscillations (ENSO MEI v.2 and AMO)  and changes in the phytoplankton community.
    Our analysis focuses exclusively on bottom-up drivers of community changes, because these data are available throughout the entire time series and at a monthly resolution. Our study builds on and expands substantially the work of \citeA{taylor_ecosystem_2012} and \citeA{pinckney_phytoplankton_2015} with a more detailed analysis of biodiversity aspects using all data available from the CARIACO time series up to the final cruise in 2017. Understanding the regional dynamics of the Cariaco Basin is crucial for unveiling theecological changes affecting this tropical marine ecosystem and the variables driving these changes.

\section{Materials and Methods}
% Here is text on Materials and Methods.
%
\subsection{Study Site}
    %Talk about the Basin shortly, then about the time series
    The Cariaco Ocean Time Series Program was established to explore the relationship between ocean surface processes and the sinking flux of particulate carbon. The Cariaco Basin (Figure~\ref{fig:map}) is a large (approximately 160 km long and 70 km wide) and deep (approximately 1,400 m) basin, situated on the northeastern Venezuelan continental shelf, with a sill to the north, running almost parallel to the coastline, at a mean depth of around \qty{100}{m}. This area is relatively productive ($\sim$ \qty{400}{g.C.m^{-2}.y^{-1}}), with primary production and vertical particulate organic matter fluxes exceeding those observed at BATS and HOT sites, resulting in the formation of anoxic conditions below \qty{\sim 200}{m} depth.

\begin{figure}
\noindent\includegraphics[width=\textwidth]{fig/Map_CAR.pdf}
\caption{Bathimetric map of the CARIACO Basin in the Caribbean Ocean, off the coast of Venezuela, indicating the CARIACO Mooring, the station of the Ocean Time-Series (located at \ang{10.50}N, \ang{64.67}W). A sill, running almost parallel to the coast at an average depth of 100\,m, from the Isla de Magarita and the Isla de la Tortuga and beyond, encloses the Cariaco basin from the North.}
\label{fig:map}
\end{figure}

\subsection{In-Situ Data}
    Samples were collected as part of the Cariaco Ocean Time Series Program by means of monthly research cruises at a station located in the eastern Cariaco Basin (\ang{10.50}N, \ang{64.66}W) \cite{muller-karger_scientific_2019}.
    The CARIACO field program consisted of a monthly core oceanographic cruise with Niskin bottle and CTD deployments, in addition to other deployments and samples not used in this work. Field logistics and partial sample analysis were coordinated by the Fundación La Salle de Ciencias Naturales on Margarita Island, Venezuela. Details about the sampling procedures adopted in CARIACO, sampling schemes, analytical methods, data quality control, and inter-calibration procedures, have been already described by \citeA{mullerkarger_annual_2001, thunell_particulate_2007, astor_yrene_m_handbook_2013, astor_interannual_2013, astor_sintesis_2014}.

@@

    Between late 1995 and early 1998, nutrient data was provided by William Senior (Universidad de Oriente, Cumana, Venezuela) and from mid-1997 to present by Dr. Kent Fanning (University of South Florida, St. Petersburg, USA). Samples from eight cruises were analyzed by both laboratories. Data for PO$_4$ and SiO$_4$ were in strong agreement (within ±5\%), whereas  NO$_\text{x}$ and NH$_4$ measurements showed discrepancies. \cite{taylor_ecosystem_2012}. We used a merged dataset for all nutrients, in which coincidental monthly measurements are averaged.

@@

    Data collected with Niskin bottles, CTD, and phytoplankton taxonomy were retrieved from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) \url{https://www.bco-dmo.org/project/2047}. 
    From the CARIACO Time Series data, we used only those variables that cover the entire length of the time series without large gaps (i.e. no more than three months) in coverage. In total, we used a complete set of 254 monthly data points, for which all variables were available.

\subsection{Climate data}
    %from Taylor et al. 2012
    Zonal wind stress (-u, easterly Trade Winds) across the southern Caribbean Sea drives nearshore upwelling of nutrient-rich waters via Ekman transport \cite{rueda-roa_southern_2013}. We extracted the -u wind component at \qty{10}{\meter} from the ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis in the grid cell closest to the Cariaco time series sampling location (\ang{10.4}N, \ang{64.8}W, 31 km grid size), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels}. For simplicity, we refer to this variable as "wind speed" throughout the article. The data for precipitation and evaporation are also extracted from the ERA5 dataset for the aforementioned grid cell. Evaporation provides an estimate of the accumulated amount of water that has evaporated from Earth's surface, while precipitation is an estimate for total precipitation, comprising all water that falls on Earth's surface as rain or snow.

    The monthly AMO index data (10-year low-pass), which represents an area-averaged, detrended low-pass filtered North Atlantic SST anomalies, were obtained from \url{https://climatedataguide.ucar.edu/climate-data/atlantic-multi-decadal-oscillation-amo}.

    The bi-monthly multivariate El Niño/Southern Oscillation (ENSO) indices (MEI v.2), defined as the time series of the leading combined Empirical Orthogonal Function (EOF) of five different variables (sea level pressure (SLP), sea surface temperature (SST), zonal and meridional components of surface wind, and outgoing long-wave radiation (OLR)) over the tropical Pacific basin (\ang{30}S-\ang{30}N and \ang{100}E-\ang{70}W), were obtained from \url{https://psl.noaa.gov/enso/mei/}. In our analysis, bi-monthly values are aligned with the tail month (December-January is treated as the value for January).

    \subsubsection{Data processing}
    % Niskin, CTD and Phytoplankton data processing
    Data were processed and analyzed using the R programming language (version 4.3.3) \cite{r_core_team_r_2024} with data pipelines created using tidyverse packages \cite{wickham_welcome_2019}. All the codes we produced for data processing and data analysis are publicly available on GitHub \url{https://github.com/ben1post/PhytoCariaco}.
    For all discrete depth measurements, the data were interpolated using the "oce" package \cite{kelley_oce_2023} with the adaptive "unesco" algorithm defined by the U.S. National Oceanographic Data Center \cite{johnson2006world}. The interpolated values were depth integrated to 100 meters and in four discrete intervals between 0-25, 25-50, 50-75 and 75-100 \meter for further analysis. Additionally, sea surface temperature (SST) was calculated from CTD temperature measurements averaged over the top \qty{10}{\meter}. For phytoplankton microscopy data, individually identified taxonomic units (species or genus) were interpolated and integrated across depth, and finally grouped by genus via the sum of interpolated counts.

\subsection{Data analysis}    
    To provide an overview of the data available  and the general trends shown by the data, we calculated annual means for relevant variables and normalized the data range as z-scores ($z = \frac{(x – \mu)}{\sigma}$). Cruises from 1995, 2016, and 2017 were excluded because insufficient coverage in these years prevented us from producing reliable yearly mean estimates.

@@

    \subsubsection{Analysis of Phytoplankton Community Structure}
    To investigate the structure of the phytoplankton community over time, we used a complete linkage clustering method via the "hclust" function in R \cite{r_core_team_r_2024} on the binary Jaccard distance matrix of annual aggregates of observed genera integrated over 100 m depth. Additionally, we performed non-metric multidimensional scaling (NMDS) ordination and hierarchical clustering on the monthly integrated counts that were square-root transformed using "metaMDS" from the "vegan" package \cite{oksanen_vegan_2024}. We employed the Bray distance matrix in a two-dimensional ordination that reached a converging solution within 30 steps. To relate the multidimensional structuring of the community data to variability in environmental data, we used the "envfit" function from the "vegan" package to fit the environmental vectors onto the ordination with 999 permutations. 
    To contrast the environmental parameters and diversity data between the community clusters resulting from the yearly aggregated clustering, we created density distribution plots contrasting the monthly measurements between clusters using the "ggpubr" package \cite{kassambara_ggpubr_2023}. To test whether the environmental variables, chlorophyll $a$ and diversity metrics varied significantly between the two community clusters, we employed a two-sided Wilcoxon sum rank test using the "wilcox.test" function in R \cite{r_core_team_r_2024}.
    % detailed explanation of NMDS in Winder and Hunter 2008
